---
title: "Customizing the visualizations of CalibraCurve"
author: 
  - name: Karin Schork
    affiliation:
    - Medizinisches Proteom-Center, Ruhr-Universität Bochum
    email: karin.schork@rub.de
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('CalibraCurve')`"
vignette: >
  %\VignetteIndexEntry{2. Customizing the visualizations of CalibraCurve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL 
## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    CalibraCurve = citation("CalibraCurve")[1]
)
```


# Introduction

```{r "start", message=FALSE}
library("CalibraCurve")

file <- system.file("extdata", "MSQC1/msqc1_dil_GGPFSDSYR.rds", 
                    package = "CalibraCurve")

D <- readDataSE(file, concColName = "amount_fmol", substColName = "Substance")

RES <- CalibraCurve(D)

```


# Handling multiple calibration curves

If multiple analytes are present in the data, `CalibraCurve` by default plots 
each calibration curve separately and saves them in separate files. 
There are, however, two possibilities to combine the calibration curves.

### Multiplot

The multiplot option allows to arrange the different curves in a grid. This 
functionality is based on the `facet_wrap` function from `ggplot2`.

```{r "multiplot", message=FALSE}
RES <- CalibraCurve(D, plot_type = "multiplot")
RES$plot_CC_list

```

The single plots are arranged in a grid. The number of rows and columns is 
chosen automatically by default, but can be changed by using the 
`multiplot_nrow` and `multiplot_ncol` parameters:

```{r "multiplot_nrow_ncol", message=FALSE, out.width="100%"}
RES <- CalibraCurve(D, plot_type = "multiplot", multiplot_nrow = 3, 
                    multiplot_ncol = 4)
RES$plot_CC_list

```


By default, the x and y-axis ranges are individual per plot 
`multiplot_scales = "fixed"`. 
Sometimes it makes sense to fix both axis or only one of them for better 
comparability. For this, the `multiplot_scales` parameter
can be set to `fixed`, `fixed_y` or `fixed_x`: 

```{r "multiplot_fixed", message=FALSE, out.width="100%"}
RES <- CalibraCurve(D, plot_type = "multiplot", multiplot_scales = "fixed")
RES$plot_CC_list

```

Now, the x- and y-axes are scaled the same for each plot, so the intensity 
values can be directly compared between plots.


### All-in-one plot

As a second option, the `"all_in_one"` option allows to plot all curves into 
the same graphic, so a more direct comparison is possible:

```{r "multiplot_allin1", message=FALSE, out.width="100%"}
RES <- CalibraCurve(D, plot_type = "all_in_one")
RES$plot_CC_list

```

Each of the 9 calibration curves is presented in the same plot in different 
colours (see legend). Please note that for better visability, the 
linear ranges are not indicated by the grey background anymore. However, data 
points outside of the linear range are still plotted with a higher transparency 
(alpha-value).


# Changing colours

Different colours can be changed in the plots. For the calibration curves, the 
colour of the data points (`point_colour`), the colour of the curve 
(`curve_colour`) and of the linear range (`linear_range_colour`) can be 
customized. The colours can be given as the colour name or the hexadecimal 
colour code:


```{r "CC_colours", message=FALSE}
RES <- CalibraCurve(D, point_colour = "blue", curve_colour = "black", 
                    linear_range_colour = "red")
RES$plot_CC_list[[1]]

```


For the response factor plots, the colour of data points and corresponding 
connection lines within and outside of the thresholds (`RF_colour_within` and 
`RF_colour_outside`) can be adapted. Furthermore, the colour of the threshold 
lines can be changed (`RF_colour_threshold`).


```{r "RF_colours", message=FALSE}
RES <- CalibraCurve(D, RF_colour_threshold = "grey", 
                    RF_colour_within = "purple", 
                    RF_colour_outside = "darkgreen")
RES$plot_RF_list[[1]]

```




# Other plot elements

The plotting of the data points can be omitted. This makes sense especially 
when multiple curves are plotted in the same plot. As a downside, information 
on the linear ranges is lost:


```{r "data_points", message=FALSE}
RES <- CalibraCurve(D, plot_type = "all_in_one", show_data_points = FALSE)
RES$plot_CC_list

```


It is also possible to remove the background that indicates the linear range. 
Also, it is possible to write the regression curve equation (upper left corner 
of the plot) and the R squared value into the plot, which is turned off by 
default:

```{r "regression_info", message=FALSE}
RES <- CalibraCurve(D, show_linear_range = FALSE, show_regression_info = TRUE)
RES$plot_CC_list[[1]]

```


Furthermore, the axis labels can be changed, in this example we specify the 
labels to match the underlying data set:


```{r "axis labels", message=FALSE}
RES <- CalibraCurve(D, xlab = "Amount (fmol)", ylab = "Area")
RES$plot_CC_list[[1]]

```




It has to be noted that the plots are returned as ggplot2 objects by the 
respective functions. Therefore, even more customizations can be done by using 
the ggplot2 functionality. For example, we can give the plot a title and use a 
different theme which does not contain a grid in the background:


```{r "ggplot_object", message=FALSE}
RES <- CalibraCurve(D)
pl <- RES$plot_CC_list[[1]]

library(ggplot2)
pl <- pl + ggtitle("Calibration Curve") + theme_classic()
pl

```






# R session information

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["BiocStyle"]])` with `r CRANpkg("knitr")` 
`r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` 
`r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r Biblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
